---
title: "Week 3 Homework"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
libraries <- c("tidyverse", "shiny", "excelR", "readxl", "knitr", "vroom")
lapply(libraries, library, character.only = TRUE)
```

# Summary 

This document is broken down into subsections corresponding to the tasks given for homework. The primary task is two write an R Markdown file which contains code that carries out certain tasks.

## Task 1

Load `to_sort_pop_1.csv` and `to_sort_pop_1.csv` from [bioinformatics_data](https://github.com/chrit88/Bioinformatics_data/tree/master/Workshop%203) provided by Dr. Chris Clements on github.

```{r include=TRUE}

tsp1 <- vroom("https://raw.githubusercontent.com/chrit88/Bioinformatics_data/master/Workshop%203/to_sort_pop_1.csv")
head(tsp1)

tsp2 <- vroom("https://raw.githubusercontent.com/chrit88/Bioinformatics_data/master/Workshop%203/to_sort_pop_2.csv")
head(tsp2)

```

## Task 2

Using a `tidyverse` function, join both of these data together into a single tibble.

```{r include=TRUE}
tsp_joined <- full_join(tsp1, tsp2)
head(tsp_joined)

```

## Task 3

Reshape the tibble from wide to long format and include 3 new columns.

```{r include=TRUE}

# take the tsp_long tibble and filter it into the pivot_longer function to convert from wide to long format

tsp_long <- tsp_joined %>%
                
# specifying the column titles for the converted tibble
  
                pivot_longer(cols = -c(species:tertiary_threat),
                             
                             # adding two new column titles under which unmatched character strings and our pop number will be
                             # assigned to
                             
                             names_to = c("population", "date"),
                             
                             # adding new column under which unmatched values will be assigned to
                             
                             values_to = "abundance",
                             
                             # using regular expressions to search for population number and date in character strings. The 
                             # brackets are the capture groups, and each group is assigned to one of the new columns.
                             names_pattern = "pop_(1|2)_(.*)",
                             )
head(tsp_long)

```
## Author

Conor Noonan