---
title: "Programming in R Final Coursework"
author: "Student Number: 2185380"
date: "10/01/2022"
output: github_document
---

```{r setup, include=TRUE, message=FALSE, warning=FALSE, results='hide'}

knitr::opts_chunk$set(echo = TRUE)

# Loading libraries
libraries <- c("tidyverse",
               "knitr",
               "vroom",
               "DHARMa",
               "fitdistrplus",
               "glmmTMB",
               "gamlr",
               "performance",
               "see",
               "insight")
lapply(libraries, library, character.only = TRUE)

```

## Reading, Manipulating, and Visualizing Data

```{r include=TRUE, results=FALSE, message=FALSE}

#Using the Vroom package due to its speed and compatability with .gz compressed files
mass_raw <- vroom("~/R/bioinformatics/tb1_exam/data/mass_pu21373.tsv.gz")

nao_raw <- vroom("~/R/bioinformatics/tb1_exam/data/NAO_pu21373.tsv.gz")

```

I first wanted to calculate the annual mean NAO and summer NAO and merge the data.

```{r include=TRUE, message=FALSE}

# Making a new dataframe containing annual sNAO
sNAO <- nao_raw %>%
  filter(month == "June" | month == "July" | month == "August") %>%  
  group_by(year) %>%
  mutate(snao = mean(NAO)) %>%
  dplyr::select(snao, year) %>%
  unique()      # Removing excess rows

# Joining the datasets
data_combined <- left_join(mass_raw, sNAO, by="year")

```

Now that all of the data was in one dataframe, I wanted to tidy it up.
  
```{r include=TRUE}

data_combined <- data_combined %>%
  na.omit(weight) %>%       # Removing rows which have no measurements for weight
  filter(weight > 0) %>%    # Removing rows which have negative weight values
  group_by(year) %>%
  mutate(mass_mean = mean(weight),    # Calculating annual mean mass for visualization
         std_time = (year - 1980),    # Converting year to a numeric vector
         site = as.factor(site))      # Converting site to a factor

head(data_combined, n=5)
```

I wanted to visualize the data to see if there were any potential relationships between weight and other variables. 

```{r include=TRUE, message = FALSE}

# Average weight measurements over time

time_plot <- ggplot(data = data_combined, aes(x=year, y=mass_mean)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Annual Mean Weight over Time",
    x = "Time (Year)",
    y = "Mean Weight (kg)") +
  geom_smooth(method = "loess") +
  theme_bw()  

# Relationship between annual weight measurements and annual snao measurements

snao_plot <- ggplot(data = data_combined, aes(x = snao, y = mass_mean)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Annual Mean Weight vs Annual sNAO",
    x = "NAO Index",
    y = "Weight (kg)") +
  geom_smooth(method = "loess") +
  theme_bw()  

# Relationship between annual weight measurements and population size

pop_plot <- ggplot(data = data_combined, aes(x=pop_size, y=mass_mean)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Annual Mean Weight vs Population",
    x = "Population",
    y = "Weight (kg)") +
  geom_smooth(method = "loess") +
  theme_bw() 

# Relationship between weight and sex 

sex_plot <- ggplot(data=data_combined, aes(x=as.factor(sex), y=weight), color =sex) +
  geom_boxplot(
    fill = c("#CC6666", "#9999CC"), 
    color = "black",
    outlier.colour="red",
    outlier.fill="red"
  ) +
  labs(
    title = "Weight vs Sex",
    x = "Sex",
    y = "Weight (kg)"
  ) +
  scale_x_discrete(labels = c("Female", "Male")) +
  theme_bw() +
  theme(legend.position = "none")

# Relationship between weight and sample site

site_plot <- ggplot(data = data_combined, 
                    aes(x = as.factor(site), y = weight, fill=as.factor(site))) +
  geom_boxplot() +
  labs(
    title = "Weight vs Site",
    x = "Site",
    y = "Weight (kg)"
  ) +
  theme_bw() +
  scale_color_brewer("Set2") +
  theme(legend.position = "none")

# Relationship between annual mean weight and number of offspring

babies_plot <- ggplot(data = data_combined, 
                      aes(x=as.factor(n_babies), y=weight, fill=as.factor(n_babies))) +
  geom_boxplot() +
  labs(
    title = "Annual Mean Weight vs Number of Offspring",
    x = "Number of Offspring",
    y = "Weight (kg)"
  ) +
  theme_bw() +
  scale_color_brewer("Set2") +
  theme(legend.position = "none")

# Relationship between weight and sex

age_plot <- ggplot(data = data_combined, 
                   aes(x = as.factor(age), y=weight, fill=as.factor(age))) +
  geom_boxplot() +
  labs(
  title = "Weight vs Age",
  x = "Age (Years)",
  y = "Weight (kg)"
  ) +
  scale_colour_steps() +
  theme_bw() +
  theme(legend.position = "none")

```

I wanted to first see, on average, how the mass of Soay sheep was trending over time, as well as its relationship with sNAO (`Figure 1`).

```{r, message = FALSE, include=TRUE}

# Set global margins
par(mar = c(4, 4, .1, .1))

```

```{r, message = FALSE, include=TRUE, fig.show="hold", out.width="50%", fig.cap="Figure 1."}

time_plot
snao_plot

```


Interesting relationships between other variables like sex, population number, and age were also seen (`Figure 2`).


```{r, message = FALSE, include=TRUE, fig.show="hold", out.width="50%", fig.cap="Figure 2."}

sex_plot
pop_plot
age_plot
babies_plot

```

I wanted to check if there was any variability across the sampling sites (`Figure 3`).

```{r, message = FALSE, fig.show="hold", out.width="75%", fig.cap="Figure 3."}

site_plot

```


## Statistical Modelling

To determine the distribution of the `weight` measurements, I used the `fitdistrplus` package. 

```{r, message = FALSE, fig.width=7, fig.cap="Figure 4."}

# Creating a fitdist object using the weight measurements
# Assuming Gaussian distribution as weight data is continuous
fit_gauss <- fitdist(data_combined$weight, 
                    distr = "norm")

plot(fit_gauss)

# Outputting some goodness-of-fit statistics
gofstat(fit_gauss)

```


The data appears to follow the Gaussian distribution, although the Q-Q plot, density distributions, and stats indicate that the data is somewhat positively skewed.


I created some generalized linear models (GLMs) and compared different transformations and error distributions.

```{r include=TRUE}

# Scaling variables with largely different scales
# Using a log(x + C) transformation for snao due to negative values
# Assuming the normal distribution
mod_gaus_1.1 <- glm(scale(weight) ~ log(snao + 1) + scale(std_time) + scale(age) + sex + 
                      scale(pop_size) + site + n_babies,
                      data = data_combined,
                      family = "gaussian") 

# Looking at different transformations
# Transforming variables to remove 0 and negative values
mod_gaus_1.2 <- glm(sqrt(weight) ~ log(snao + 1) + sqrt(std_time) + sqrt(age) + sex + 
                      sqrt(pop_size) + site + n_babies,
                      data = data_combined,
                      family = gaussian(link="log")) 

mod_gaus_1.3 <- update(mod_gaus_1.2, family = gaussian(link="inverse"))


# Comparing model fits using Akaike information criterion (AIC).
# The `gamlr` package  provides useful functions for calculating AIC
AIC_gaus_mods_v1 <- data.frame(model = c("mod_gaus_1.1", "mod_gaus_1.2", "mod_gaus_1.3"),
                          AIC = c(AIC(mod_gaus_1.1), AIC(mod_gaus_1.2), AIC(mod_gaus_1.3)))

AIC_gaus_mods_v1[order(AIC_gaus_mods_v1$AIC),]

```

```{r include=TRUE}

# Checking the diagnostic plots we can see that the log and inverse transformations don't 
# improve the model fit or normality
par(mfrow = c(1, 3))

plot(mod_gaus_1.1, which=2)
plot(mod_gaus_1.2, which=2)
plot(mod_gaus_1.3, which=2)

```
```{r include=TRUE}

# Testing gamma distributions due to skewness
# Transforming to make variables non-negative or 0
mod_gam_1.1 <- glm(sqrt(weight) ~ log(snao + 1) + sqrt(std_time) + scale(age) + sex + 
                     scale(pop_size) + site + n_babies,
                     data = data_combined,
                     family = Gamma(link = "log")) 

mod_gam_1.2 <- update(mod_gam_1.1, family=Gamma(link="identity"))

mod_gam_1.3 <- update(mod_gam_1.1, family=Gamma(link="inverse"))

# Testing inverse Gaussian distribution due to skewness
# Transforming to make variables non-negative or 0
mod_ingam_1.1 <- glm(sqrt(weight) ~ log(snao + 1) + sqrt(std_time) + scale(age) + sex + 
                       scale(pop_size) + site + n_babies,
                       data = data_combined,
                       family = inverse.gaussian())

mod_ingam_1.2 <- update(mod_ingam_1.1, family=inverse.gaussian(link="identity"))

mod_ingam_1.3 <- update(mod_ingam_1.1, family=inverse.gaussian(link="inverse"))

```
    
```{r include=TRUE}

par(mfrow = c(1, 3))

# Comparing Gamma Distribution
plot(mod_gam_1.1, which=2)
plot(mod_gam_1.2, which=2)
plot(mod_gam_1.2, which=2)

# Comparing inverse Gaussian Distribution
plot(mod_ingam_1.1, which=2)
plot(mod_ingam_1.2, which=2)
plot(mod_ingam_1.3, which=2)

```


From these results, the best fit to the data was the Gaussian distribution with an identity link. 


I looked at testing potential interactions and random effects.

```{r include=TRUE}

# Looking at an interaction between sNAO and std_time
mod_gaus_1.4 <- glm(scale(weight) ~ log(snao + 1)*scale(std_time) + scale(age) + sex + 
                      scale(pop_size) + site + n_babies,
                      data = data_combined,
                      family = "gaussian")

# To assess if ID should be nested within site, each ID should be unique to one site
nrow(data_combined %>%
  ungroup() %>%
  dplyr::select(ID, site) %>%
  unique())

# Number of rows is almost exactly the same as the largest ID number, meaning that
# it should be nested within site.
# Using glmmTMB to model generalized linear mixed-effects model
mod_gaus_1.5 <- glmmTMB(as.vector(scale(weight)) ~ log(snao + 1)*scale(std_time) + 
                          scale(age) + sex + scale(pop_size) + (1|site/ID),
                          data = data_combined,
                          family = "gaussian")



AIC_gaus_mods_v2 <- data.frame(model = c("mod_gaus_1.1", "mod_gaus_1.4", "mod_gaus_1.5"),
                          AIC = c(AIC(mod_gaus_1.1), AIC(mod_gaus_1.4), AIC(mod_gaus_1.5)))

AIC_gaus_mods_v2[order(AIC_gaus_mods_v2$AIC),]

```

`mod_gaus_1.5` appears to be the best fit, however I noticed that the relationship between sNAO and weight did not follow the trend visualized in `Figure 1`. The results of the model suggest that sNAO declines as weight increases.

```{r include=TRUE}

summary(mod_gaus_1.5)

```

As both `std_time` and `snao` exhibit a similar relationship with weight (`Figure 1`), these two variables may be linearly related, resulting in collinearity which may explain the poor significance and reversed trend. 

```{r include=TRUE, fig.width=10}

# CHecking for collinearity using the 'performance' package to calculate VIF
collinearity_results <- check_collinearity(mod_gaus_1.5)

collinearity_results
plot(collinearity_results)

```

The results show that there is moderate correlation seen for `std_time`. To remedy this, I removed `std_time` from the model.

```{r include=TRUE}

# Removing std_time from the model 
# Removing n_babes variable due to lack of significance or effect
mod_gaus_1.6 <- glmmTMB(as.vector(scale(weight)) ~ log(snao + 1) + scale(age) + sex + 
                          scale(pop_size) + (1|site/ID),
                          data = data_combined,
                          family = "gaussian")

summary(mod_gaus_1.6)

```

I assessed the fit of the new model and tested for deviation, heteroskedasticity, and outliers, using the `DHARMa` package.

```{r include=TRUE, message=FALSE, results='hide', warning=FALSE}

set.seed(666)

# Increasing the number of simulations to 5000 to stabilize the simulated values and
# decrease the number of random outliers. 
# Error message pops up as we're using a GLM instead of a GLMM as input, but this 
# has little impact on the deviation statistics or outliers.
mod_gaus_1.6_sim <- simulateResiduals(mod_gaus_1.6, n = 5000)

# Testing for outliers
testOutliers(mod_gaus_1.6_sim, plot = TRUE)

```

Although there appear to be some outliers, the frequency is very low and insignificant. These may just be random and arise from too-low a number of simulations.

```{r include=TRUE, message=FALSE, results='hide', warning=FALSE}

# Looking at the residuals to see if they follow the normal distribution and to see if
# there's any significant dispersion
plotQQunif(mod_gaus_1.6_sim)

```

There is no significant deviation detected, suggesting the fit to the Normal distribution is appropriate. 


```{r include=TRUE}

# Calculating a pseudo r2 to get a very rough idea of the goodness-of-fit
# Using Nakagawa's pseudo r2 from'performance' package as we are calculating r2 for a GLMM
r2_nakagawa(mod_gaus_1.6)

```

0.8 and 0.9 are conditional and marginal values respectively.

```{r include=TRUE, fig.width=10, fig.height= 15}

check_model(mod_gaus_1.6)

```

The data appears to be homoscedastic with no significant VIF between predictors.


For some final visual checks, I plotted the predicted values to the observed values.

```{r include=TRUE}

# Generating predictions from the model
data_combined$predicted <- predict(mod_gaus_1.6,
                                   data = data_combined,
                                   predict.all=TRUE,
                                   type = "response")

#Plotting the predicted results against the observed data
mod_plot <- ggplot(data_combined, aes(x = scale(weight), 
                           y = predicted)) + 
  geom_point(col="grey") + 
  geom_abline(slope = 1, size=2) +
  theme_minimal() +
  labs(
    x = "Observed",
    y = "Predicted",
    title = "Predivded vs Observed Values"
  )

mod_plot

```


## Results

The following assumptions were made based on the summary below:

Predictor         | Relationship with Soay Sheep mass                                 | Evidence
------------------|-------------------------------------------------------------------|----------
Age               |An increase in age is associated with an increase in sheep mass    |p value ~ 0, Visual trend 
Sex               |Male sheep are on average associated with a decrease in mass compared to females |p value ~ 0, Visual trend
Population Size   |Population increase is associated with a decrease in sheep mass   |p value ~ 0, Visual trend 

The variables with the strongest influence on Soay sheep mass are age and sex as evidenced by their relatively large coefficients of 0.774390 and -0.886717 respectively. Although there is a clear correlation between `std_time` and `snao` with `weight` individually, it appears that together with other predictors their effects become 'insignificant'. 

```{r include=TRUE}

summary(mod_gaus_1.6)

```

## Data Sources

NAO and Soay sheep mass data was provided by Dr. Chris Clements on [github](https://github.com/chrit88/R-coursework-data/tree/main/2021).
