---
title: "Week 6 Homework"
output: github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
libraries <- c("tidyverse", "knitr", "vroom", "lubridate","devtools", "patchwork", "multcomp", "countrycode", "wbstats")
lapply(libraries, library, character.only = TRUE)
install_github("thomasp85/patchwork")

```

# Homework Part 1 - Tokyo Olympics

For the first part of today's homework, I'll be looking at whether there is any significant effect of a country’s GDP on their position in the Tokyo Olympics medal table. The data used was obtained from GitHub user `chrit88` [here]("https://github.com/chrit88/Bioinformatics_data/tree/master/Workshop%205").


## Task 1

Find out how to get GDP for the countries. Firstly, let's import the dataset.

```{r include=TRUE}

tokyo <- vroom("https://raw.githubusercontent.com/chrit88/Bioinformatics_data/master/Workshop%205/Tokyo%202021%20medals.csv")

## adding country codes

tokyo$code <- countrycode(tokyo$Country,
  origin = "country.name",
  destination = "iso3c")

## manually adding China

tokyo$code[2] <- "CHN"

head(tokyo)
```

The World Bank provides data on GDP which can be accessed through the wbstats package. Let's get this data and filter to include the latest data (2020).

```{r include=TRUE}

gdp <-  wb_data("NY.GDP.MKTP.CD")

gdp <- gdp %>%
  filter(date == "2020") %>%
  rename("code" = "iso3c", "gdp" = "NY.GDP.MKTP.CD")

```

## Task 2

Now that the data is loaded in, let's merge the tables together based on the country code.

```{r include=TRUE}

tokyo_gdp <- left_join(tokyo, gdp, by="code")

head(tokyo_gdp)
```

China was manually changed as it was recognized as Taiwan in the Olympic Table. 

In order to compare GDP to position in the table, we're going to make a column which contains the indexes of the rows.

```{r include=TRUE}

tokyo_gdp$position <- as.numeric(rownames(tokyo_gdp))

## remove rows that GDP couldn't be obtained for

tokyo_gdp <- tokyo_gdp %>%
  filter(!is.na(gdp))

```

## Task 3

We now want to visualise the relationship between GDP and position in the table.

```{r include=TRUE}

p1.1 <- ggplot(tokyo_gdp, aes(x =gdp, y = position)) +
  geom_point() +
  ggtitle("Relationship between GDP and Position in the Olympic Table") +
  theme_bw() +
  scale_y_reverse(breaks = c(1, 10, 20, 30, 40, 50, 60, 70, 80, 90)) +
  scale_x_reverse() 

## Makeing a plot using the log scale

p1.2 <- ggplot(tokyo_gdp, aes(x =log(gdp), y = position)) +
  geom_point() +
  ggtitle("Relationship between log(GDP) and Position in the Olympic Table") +
  theme_bw() +
  scale_y_reverse(breaks = c(1, 10, 20, 30, 40, 50, 60, 70, 80, 90)) +
  scale_x_reverse() 

p1.1 + p1.2
```

## Task 4

Fit an appropriate GLM to the data and check its assumptions.

```{r include=TRUE}

#===================================================
#         MOD1
#===================================================
mod1 <- glm(position ~ log(gdp),
            data = tokyo_gdp,
            family = "gaussian")

tokyo_gdp$pred_gaussian <- predict(mod1,
                                    type="response")

tokyo_gdp$resid_gaussian <- resid(mod1)

##Checking the distribution of the residuals to see if they fit the normal distribution

plot_mod1 <- ggplot(tokyo_gdp, aes(x = resid_gaussian)) +
  geom_histogram() +
  theme_minimal() +
  ggtitle("Histogram of residuals (gaussian with identity link)")

plot_mod1
#===================================================
#         MOD2
#===================================================

mod2 <- glm(position ~ gdp,
            data = tokyo_gdp,
            family = gaussian(link = "log"))

tokyo_gdp$pred_gaussian_log <- predict(mod2,
                                    type="response")

tokyo_gdp$resid_gaussian_log <- resid(mod2)

##Checking the distribution of the residuals to see if they fit the normal distribution

plot_mod2 <- ggplot(tokyo_gdp, aes(x = resid_gaussian_log)) +
  geom_histogram() +
  theme_minimal() +
  ggtitle("Histogram of residuals (gaussian with log link)")

plot_mod2


##compare the models
AIC_mods <- AIC(mod1,
                mod2)

## rank them by AIC using the order() function
AIC_mods[order(AIC_mods$AIC),]

```

According to the AIC criterion, there appears to be no significant difference between the two models. The residuals all appear to follow a normal distribution.

Lets graph the two models.

```{r include=TRUE}

#===========================
#     Plot for Model 1
#===========================


p2 <- ggplot(tokyo_gdp, aes(x = log(gdp), y = position)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  ylab("Position") +
  xlab("Log GDP") +
  scale_y_reverse(breaks = c(1, 10, 20, 30, 40, 50, 60, 70, 80, 90)) +
  scale_x_reverse() 

##add in a line of the predicted values from the model
p2 <- p2 + geom_smooth(data=tokyo_gdp,
                       method="glm",
                       method.args = list(family = "gaussian"),
                       formula = y ~ x,
                       col = "dodgerblue",
                       fill = "lightblue")

p2 <- p2 +geom_segment(aes(xend= log(gdp), yend = pred_gaussian),
                       col = "red")

## add a title
p2 <- p2 + ggtitle("Fitted Model (Gaussian with Ddentity Link)")


#===========================
#     Plot for Model 2
#===========================


p3 <- ggplot(tokyo_gdp, aes(x = gdp, y = position)) +
  geom_point()+
  ylab("Position") +
  xlab("GDP") 

p3 <- p3 + geom_smooth(data=tokyo_gdp,
                       method="glm",
                       method.args = list(family = gaussian(link="log")),
                       formula = y ~ x,
                       col = "dodgerblue",
                       fill = "lightblue") +
                        ggtitle("Fitted Model (With Log Link)") 

## Comparing the two

p2 + p3

```
```{r include=TRUE}
summary(mod1)

summary(mod2)
```

Looking at the summary tables and plots, it looks like there is a significant relationship between GDP and log(GDP) against position on the Olympic Table. Both relationships show that there is a negative relationship between the two, so as you decrease GDP there is a corresponding decrease in your position on the table.

------------------------------------------------------------------------------------------------------------------------------------

# Homework Part 2 - Iris with Multiple Predictors

## Task 1

Load in the Iris data and visualize the relationship between petal length and petal width.

```{r include=TRUE}

## visualize the data

irplot <- ggplot(data = iris, aes(x= Petal.Length, y=Petal.Width, col=Species)) +
  geom_point() +
  xlab("Petal Length") +
  ylab("Petal WIdth") 



## see if data conforms to gaussian

gaus_plot <- ggplot(iris, aes(x=Petal.Length, fill=Species)) +
  geom_histogram(binwidth=.1, alpha=.5, position="identity")

irplot + gaus_plot
```

## Tasks 2 - 5

The data appears to have a strong positive relationship between the independent and dependent variables. The data also conforms to the normal distribution. Because of this, I'm going to use a GLM and assume the residuals follow a normal distribution. 

```{r include=TRUE}

#==========================================
#       MOD1
#==========================================

## make a model

irmod1 <- glm(Petal.Width ~ Petal.Length,
            data = iris,
            family = "gaussian")

## calculate the residuals

iris$pred_gaussian <- predict(irmod1,
                                   type="response")

iris$resid_gaussian <- resid(irmod1)


## plot the residuals

resid_plot1 <- ggplot(iris, aes(x = resid_gaussian)) +
  geom_histogram(fill="goldenrod") +
  theme_minimal() +
  ggtitle("Histogram of residuals")

## plot the model

irplot1 <- irplot + geom_smooth(data=iris,
                       method="glm",
                       method.args = list(family = "gaussian"),
                       formula = y ~ x,
                       col = "dodgerblue",
                       fill = "lightblue")
irplot1

summary(irmod1)
```

This model appears to fit well. The residuals follow a normal distribution, the diagnostic plots don't raise any flags, and there is a strong relationship between the two variables.

## Task 6

Now lets see if the effect of petal length on petal width varies across the different species.

```{r include=TRUE}

irmod2 <- glm(Petal.Width ~ Petal.Length*Species,
            data = iris,
            family = "gaussian")

irmod3 <- glm(Petal.Width ~ Petal.Length*Species,
            data = iris,
            family = gaussian(link = "inverse"))


##compare the models
AIC_mods <- AIC(irmod2,
                irmod3)

## rank them by AIC using the order() function
AIC_mods[order(AIC_mods$AIC),]

## plotting three different lines

iris_split <- ggplot(iris, aes(x=Petal.Length, y=Petal.Width, col=Species)) +
  geom_point() +
  geom_smooth(method = "glm",
              method.args = list(family = "gaussian"),
              formula = y ~ x)+
              xlab("Petal Length") +
              ylab("Petal WIdth")

iris_split

summary(irmod2)
```

